---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gsaot

<!-- badges: start -->
[![R-CMD-check](https://github.com/pietrocipolla/gsaot/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pietrocipolla/gsaot/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The package `gsaot` provides a set of tools to compute and plot Optimal Transport (OT) based sensitivity indices. The core functions of the package are:
- `ot_indices()`: compute OT indices for multivariate outputs using different solvers for OT (network simplex, Sinkhorn, and so on).
- `ot_indices_wb()`: compute OT indices for univariate or multivariate outputs using the Wasserstein-Bures semi-metric.
- `ot_indices_1d()`: compute OT indices for univariate outputs using OT solution in one dimension.
The package provides also functions to plot the resulting indices and the inner statistics.

## Installation

You can install the development version of gsaot from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pietrocipolla/gsaot")
```

## Example

A basic application of the functions implemented in the package:

```{r example}
library(gsaot)

N <- 1000

# Define the inputs distribution
mx <- c(1, 1, 1)
Sigmax <- matrix(data = c(1, 0.5, 0.5, 0.5, 1, 0.5, 0.5, 0.5, 1), nrow = 3)

# Sample the inputs
x1 <- rnorm(N)
x2 <- rnorm(N)
x3 <- rnorm(N)

x <- cbind(x1, x2, x3)
x <- mx + x %*% chol(Sigmax)

# Define the (linear)  model and the output
A <- matrix(data = c(4, -2, 1, 2, 5, -1), nrow = 2, byrow = TRUE)
y <- t(A %*% t(x))

x <- data.frame(x)

M <- 25

# Compute the sensitivity indices using Sinkhorn's solver and default parameters
sensitivity_indices <- ot_indices(x, y, M)
sensitivity_indices

# Compute the sensitivity indices using the Network Simplex solver and default parameters
sensitivity_indices <- ot_indices(x, y, M, solver = "wasserstein")
sensitivity_indices

# Compute the Wasserstein-Bures indices
sensitivity_indices <- ot_indices_wb(x, y, M)
sensitivity_indices

# Compute the sensitivity map using 1-dimensional solver
sensitivity_indices <- ot_indices_smap(x, y, M)
sensitivity_indices
```
