---
title: "Pick an OT solver"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pick an Optimal Transport solver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
When first navigating the world of Optimal Transport, a reasonable question may be: how should I solve this optimization problem? This vignette provides a practical guide to selecting the right solver and function in `gsaot`, based on output dimensionality, data size, and the kind of information you want from the indices. The guide is intended for users who are new to OT-based sensitivity analysis and want a clear starting point for their analyses. 
```{r setup}
library(gsaot)
```

## A quick decision guide

Use the following three dimensions to guide your choice:

1. **Output dimensionality**
   - If the output is one-dimensional, use `ot_indices_1d()` or `ot_indices_wb()`.
   - If the output is multi-dimensional, use `ot_indices()` or `ot_indices_wb()`.

2. **Data size and bootstrapping**
   - For small to medium data sets (roughly up to a few thousand points), `ot_indices()` with `solver = "transport"` is a safe and accurate choice.
   - For large data sets, or when bootstrapping is needed, prefer `ot_indices()` with `solver = "sinkhorn"` to reduce computational cost.

3. **Type of information desired**
   - If you want the Wasserstein-Bures decomposition (advective and diffusive components), use `ot_indices_wb()`.

The next sections expand on each dimension and provide minimal examples.

## Dimension 1: output is 1D vs multi-D

If the output is one-dimensional, `ot_indices_1d()` is specialized and faster. It avoids building large cost matrices and is the recommended default for scalar outputs.

```{r}
set.seed(1)
N <- 800

x <- data.frame(
  x1 = rnorm(N),
  x2 = rnorm(N)
)
y1 <- 2 * x$x1 - x$x2 + rnorm(N)

res_1d <- ot_indices_1d(x, y1, M = 20)
res_1d
```

For multi-dimensional outputs, use `ot_indices()` for the standard OT indices or `ot_indices_wb()` if you need the Wasserstein-Bures information.

```{r}
Y <- cbind(y1, 0.5 * x$x1 + rnorm(N))

res_md <- ot_indices(x, Y, M = 20)
res_md
```

## Dimension 2: solver choice for `ot_indices()`

The function `ot_indices()` supports three solvers:

- `solver = "transport"`: solves the non-regularized OT problem via the `transport` package. This is a solid default for small to medium data sets when you want a close match to the exact OT solution.
- `solver = "sinkhorn"`: solves the entropic-regularized OT problem. This is faster and more scalable, especially when you want bootstrapping.
- `solver = "sinkhorn_log"`: numerically more stable for very small regularization, at a higher cost.

### Small to medium data: `solver = "transport"`

```{r}
res_transport <- ot_indices(
  x, Y, M = 20,
  solver = "transport",
  solver_optns = list(method = "networkflow")
)
res_transport
```

### Large data or bootstrapping: `solver = "sinkhorn"`

For big experiments or bootstrap confidence intervals, `solver = "sinkhorn"` is usually the best trade-off.

```{r}
res_sinkhorn <- ot_indices(
  x, Y, M = 20,
  solver = "sinkhorn",
  solver_optns = list(epsilon = 0.05, numIterations = 1000),
  boot = TRUE, R = 200
)
res_sinkhorn
```

### Regularization and the speed-accuracy trade-off

The main knob for Sinkhorn is the regularization parameter `epsilon`:

- Larger `epsilon` yields faster convergence but more smoothing, so indices may be less sensitive to fine structure.
- Smaller `epsilon` yields solutions closer to non-regularized OT but is slower and may require `sinkhorn_log` for stability.

You can explore this trade-off by running the same problem with different `epsilon` values.

```{r}
res_eps_fast <- ot_indices(
  x, Y, M = 20,
  solver = "sinkhorn",
  solver_optns = list(epsilon = 0.1)
)

res_eps_precise <- ot_indices(
  x, Y, M = 20,
  solver = "sinkhorn_log",
  solver_optns = list(epsilon = 0.005, numIterations = 2000)
)

res_eps_fast
res_eps_precise
```

## Dimension 3: Wasserstein-Bures indices

When you want more structure in the indices, `ot_indices_wb()` provides a decomposition into advective and diffusive components. This is useful when the goal is not only ranking inputs but also understanding how the output distribution changes. The function is defined for any output dimensionality.

```{r}
res_wb <- ot_indices_wb(x, Y, M = 20)
res_wb
```

## Final checklist

- Output is 1D: `ot_indices_1d()` or `ot_indices_wb()`
- Output is multi-D and you want standard indices: `ot_indices()`
- Output is multi-D and you want Wasserstein-Bures components: `ot_indices_wb()`
- Small to medium data: `ot_indices(..., solver = "transport")`
- Large data or bootstrap: `ot_indices(..., solver = "sinkhorn")`

This workflow matches the intended use of the solvers in `gsaot` and should cover most practical scenarios.
